```{r introduction-setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "figures/introduction-")
library(ggplot2)

transparent_theme()
```

# Introduction

a bit about the importance of models, AI, assistive, 

## Types of Models

descriptive, predictive, inferential

regression, classification

## Some terminology

supervise, unsupervised

types of predictors


## Where does modeling fit into the scientific process? 

```{r data-science-model, echo = FALSE, out.width = '80%', fig.cap = "The data science process (from _R for Data Science_)."}
knitr::include_graphics("figures-premade/data-science-model.svg")
```


## Modeling is a _process_, not a single activity


```{r modeling-process, echo = FALSE, out.width = '100%', fig.width=8, fig.height=3, dev = "svg", fig.cap = "A schematic for the typical modeling process.", warning = FALSE}
widths <- c(8, 4, 10, 2, 6, 6, 
            rep(1, 19), 2,
            rep(1, 19), 2,
            rep(1, 19), 2,
            rep(1, 19), 2,
            4, 8, 15,
            rep(1, 29), 6,
            rep(1, 29), 4,
            1)
srt <- c(1, cumsum(widths))
stp <- srt[-1]
srt <- srt[-length(srt)]

diag_cols <- c(EDA = "#377EB8", "Quantitative Analysis" = "#A6CEE3", 
               "Feature Engineering" = "#4DAF4A", "Model Fit" = "#E41A1C", 
               "Model Tuning" = "lightgrey")

bar_loc <- data.frame(srt = srt,
                  stp = stp,
                  g = c("EDA", "Quantitative Analysis", "EDA", "Quantitative Analysis", "EDA", "Feature Engineering", 
                        rep(c("Model Fit", "Model Tuning"), 40),
                        "Quantitative Analysis", "EDA", "Feature Engineering",
                        rep(c("Model Fit", "Model Tuning"), 14), "Model Fit", "Feature Engineering",
                        rep(c("Model Fit", "Model Tuning"), 14), "Model Fit", "Quantitative Analysis",
                        "Model Fit"))
bar_loc$ytop <- 1.9
bar_loc$ybot <- 1
bar_loc$g <- factor(as.character(bar_loc$g), 
                levels = c("EDA", "Quantitative Analysis", "Feature Engineering",
                           "Model Fit", "Model Tuning"))
text_loc <- data.frame(x = c(1, 8, 30, 36, 120, 124, 132, 147, 211, 215)+1,
                       y = 2.1)
text_loc$label <- letters[1:nrow(text_loc)]

mod_loc <- data.frame(x = c(45, 66, 87, 107, 162, 195)+1,
                      y = .75, 
                      label = c("Model\n#1", "Model\n#2", "Model\n#3", "Model\n#4",
                                "Model\n#2", "Model\n#4"))

ggplot(bar_loc) + 
  geom_rect(aes(fill = g, xmin = srt, xmax = stp,
                ymin = ybot, ymax = ytop), alpha = .7)  + 
  theme(legend.position = "bottom",
        axis.line=element_blank(),axis.text.x=element_blank(),
        axis.text.y=element_blank(),axis.ticks=element_blank(),
        axis.title.x=element_text(hjust = .05),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),plot.background=element_blank()) +
  scale_fill_manual(values = diag_cols, name = "") +
  geom_text(data = text_loc, aes(x = x, y = y, label = label)) + 
  geom_text(data = mod_loc, aes(x = x, y = y, label = label), size = 3) +   
  xlab("Time") + 
  ylim(c(.5, 2.25))
```



## What does it mean to be "tidy"


## Outline of future chapters

